Measurement of a dwarf galaxy rotational velocity from its Lyman-alpha line emission

### 2016-06-09 paper/letter.tex
typos

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
We explain these features by two different models: an
homogeneous sphere undergoing gas bulk rotation and an interstellar
medium composed by clumps with random motions.
It is the first time that an observed \lya\ spectrum can be explained
with these physical conditions.
We  find that both models independently require high velocities that
translate into a dynamical mass at least $10$ times larger than the
neutral mass hydrogen in the galaxy. 
We argue that a possibility to explain the high
dynamical mass is the presence of a super-massive black hole.
The implications of our findings for the study of LAEs, including the
first generation of galaxies are manifold. 
It demonstrates the importance of considering rotation and multiphase
physics under the possible conditions shaping the \lya\ spectra of the
first galaxies. 
Aditionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black holes. 
}

@@

This is the case of a multiphase interstellar medium.
The model is more complex as it must take into account the behaviour
of neutral dense hydrogen clumps dispersed in a tenous intraclump medium.  
Corresponding to the elaborate physical conditions in the model,
there are a wide variety of possible emerging spectra.
This variety includes single, double and triple peaked
spectra with different degrees of symmetry around the line's center.

@@

Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit models from the full radiative transfer
simulations for a rotating sphere (continous line) and a clumpy ISM
(dashed line).

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+17}_{-25}$ \kms,
the clumps ouflowing velocity $v_{\infty, \rm{cl}}=79^{+167}_{60}$
\kms and the fraction of the \lya emission that is comming comming from the cold clumps
$P_{cl}=0.72^{+0.20}_{-0.32}$. 
The muliphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $3.0^{+1.5}_{-1.8}\times
10^9$ M$_{\odot}$.

@@

This is evidence for a dynamical mass at least 11 (multiphase case) to
17 (rotation case) times higher than the mass in neutral hydrogen.  
There are three main possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass in old stars hass been constrained by non-detections
in the K-band to be $<XXX$M$_{\odot}$.
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

@@

These findings demonstrate the importance of including rotation and multiphase
conditions as features to model the \lya\ line in high redshift
galaxies.
Aditionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected black hole
in a dwarf galaxy, providing a new way to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.

@@

 For an homogeneous sphere the HI optical depth from its can be
 written as $\tau = \sigma_0 n D/2$, where $\sigma_0=5.898\times
10^{-14}$cm$^{-2}$ is the Lyman$\alpha$  optical depth at the 
line's center, $n$ is the number density and $D$ is the sphere's
diameter. 
From this we can impose additional constrains on $D$ from the tipical
values of the Hydrogen number density and our constrain on
$\tau=10^{7}$.  Using a range of $1<n/\mathrm{atoms/cm}^{-3} <
10^{-3}$. This gives us a range of $0.11 < D/\mathrm{kpc}<100$. 
Together from the total HI mass we have thus that the HI region should
have a diameter of $0.11 < D/\mathrm{kpc}<0.34$.

@@

The almost complete absence of outflows is a central requirement for
reproducing the lyman alpha line in \tol. In this section we present
some results on the exploration of the presence of outflows join with
the bulk rotation. We defer a detailed exploration of this effects for
another publication.

>>>>

\begin{document}
\pagestyle{empty}
\noindent
\textbf{Þú ert jörðin}
\\
\\
JEFR$^{1}$, MCRG$^1$, JNGC$^2$, MG$^3$, MD$^3$
\\
\\
\scriptsize
{$^1$ Bogota
\\
$^2$ Tucson
\\
$^3$ Oslo
\normalsize
\\
\\
\textbf{
  Star-forming Compact Dwarf Galaxies (CDGs) resemble the expected
  pristine conditions of the first galaxies in the Universe.    
Before the observational detection of the first galaxies becomes
reality, CDGs are the best systems to test our ideas on primordial
galaxy formation and evolution.    
Here we report on one of such CDGs, \tol, which presents
a broad symmetric Lyman-$\alpha$ emission that had evaded theoretical
interpretation so far. 
We explain these features by two different models: an
homogeneous sphere undergoing gas bulk rotation and an interstellar
medium composed by clumps with random motions.
It is the first time that an observed \lya\ spectrum can be explained
with these physical conditions.
We  find that both models independently require high velocities that
translate into a dynamical mass at least $10$ times larger than the
neutral mass hydrogen in the galaxy. 
We argue that a possibility to explain the high
dynamical mass is the presence of a super-massive black hole.
The implications of our findings for the study of LAEs, including the
first generation of galaxies are manifold. 
It demonstrates the importance of considering rotation and multiphase
physics under the possible conditions shaping the \lya\ spectra of the
first galaxies. 
Additionally, if future kinematic maps of \tol\ confirm the high
velocities postulated in our model, it would provide new
evidence for dwarf galaxies as hosts of supermassive black holes. 
}

@@

This is the case of a multiphase interstellar medium.
The model is more complex as it must take into account the behavior
of neutral dense hydrogen clumps dispersed in a tenuous intraclump medium.  
Corresponding to the elaborate physical conditions in the model,
there are a wide variety of possible emerging spectra.
This variety includes single, double and triple peaked
spectra with different degrees of symmetry around the line's center.

@@

Figure 1. summarizes our findings.
Dots represent the observational data for \tol with the
overplot from our best fit models from the full radiative transfer
simulations for a rotating sphere (continuous line) and a clumpy ISM
(dashed line).

@@

In the multiphase model the best constrained parameters are
the clump velocity dispersion  $\sigma_{\rm{cl}}=71^{+17}_{-25}$ \kms,
the clumps outflowing velocity $v_{\infty, \rm{cl}}=79^{+167}_{-60}$
\kms and the fraction of the \lya emission that is  coming
from the cold clumps  
$P_{cl}=0.72^{+0.20}_{-0.32}$. 
The multiphase model assumes that the clouds are distributed over a
sphere of $5$kpc in radius, close to the $\approx 4$kpc physical size
of \tol\ as determined by optical imaging.
The assumed physical size and the velocity dispersion $\sigma_{\rm
  cl}$ correspond to a  dynamical mass of  $3.0^{+1.5}_{-1.8}\times
10^9$ M$_{\odot}$.

@@

This is evidence for a dynamical mass at least 11 (multiphase case) to
17 (rotation case) times higher than the mass in neutral hydrogen.  
There are three main possibilities to explain the dynamical 
mass: stars, dark matter and a supermassive black hole. 
The stellar mass in old stars has been constrained by non-detections
in the K-band to be $<XXX$M$_{\odot}$.
Considering a standard dark matter halo, it can only contribute to
$xx$ M$_{\odot}$ in the region of interest. 
A super massive black hole remains an open possibility. 
Recent observations of ultra-compact dwarf galaxies
\cite{Seth2014} have confirmed the presence of super-massive black
holes containing $15\%$ of the total object mass, suggesting that
there is a large population of undetected black holes in dwarf
galaxies.

@@

These findings demonstrate the importance of including rotation and multiphase
conditions as features to model the \lya\ line in high redshift
galaxies.
Additionally, if the hypothesis of a supermassive black
hole in \tol\ proves to be consistent with future observational
kinematic maps, it could correspond to a so far undetected black hole
in a dwarf galaxy, providing a new way to test and probe
theories on the co-evolution of galaxies and black holes in the first
generation of galaxies.

@@

 For an homogeneous sphere the HI optical depth from its can be
 written as $\tau = \sigma_0 n D/2$, where $\sigma_0=5.898\times
10^{-14}$cm$^{-2}$ is the Lyman$\alpha$  optical depth at the 
line's center, $n$ is the number density and $D$ is the sphere's
diameter. 
From this we can impose additional constrains on $D$ from the typical
values of the Hydrogen number density and our constrain on
$\tau=10^{7}$.  Using a range of $1<n/\mathrm{atoms/cm}^{-3} <
10^{-3}$. This gives us a range of $0.11 < D/\mathrm{kpc}<100$. 
Together from the total HI mass we have thus that the HI region should
have a diameter of $0.11 < D/\mathrm{kpc}<0.34$.

@@

The almost complete absence of outflows is a central requirement for
reproducing the \lya\ alpha line in \tol. In this section we present
some results on the exploration of the presence of outflows join with
the bulk rotation. We defer a detailed exploration of this effects for
another publication.
